---
title: "Data visualization assignment"
author: "Amelia Xu"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE, fig.width = 6, fig.height = 4)
```

In this exercise, you will practice basic plotting with **`ggplot2`** on
a dataset of median gene expression levels across samples from various
human tissues from the GTEx Project. The dataset (`dt`) has one row per
gene, with columns:

- `Name`: the Ensembl gene ID
- `Description`: the gene symbol
- One column per tissue containing the median expression levels (in TPM
  units)

Insert your code into the empty code blocks. Insert answers to any
comprehension questions as comments within the code blocks. You do not
need to submit the figures themselves, only the code used to produce
them. Remember to label your axes!

**Using HopGPT.** As in the last assignment: you are welcome to use it,
and wherever you do, add a couple of comment lines in that same code
block noting what you asked it and how you checked the answer.

------------------------------------------------------------------------

### Load data

Start by loading the data into R, as before.

```{r load-data}
library(tidyverse)

dt <- read_delim("/Users/cmdb/Data/GTEx/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz",
                 skip = 2, col_names = TRUE, delim = "\t")
```

Recreate the `is_mitochondrial` column from the last assignment; you
will need it in Q4 and Q5.

```{r add-mito}
dt <- dt %>%
  mutate(is_mitochondrial = str_starts(Description, "MT-"))

dt %>% count(is_mitochondrial)
```

------------------------------------------------------------------------

### Q1. Histogram

Make a histogram of gene expression in the liver. Use the argument
`breaks = seq(0, 100, 2)` to bin in units of 2 TPM, ranging from 0 to
100. Note that this cuts off the data at 100, which might be misleading
depending on what we are trying to depict! How many genes are you
leaving out?

What do you notice about the shape of the distribution? What does it
tell you about the relative expression of genes in the liver
transcriptome? Do you think this is specific to the liver? Check another
tissue, as well.

```{r q1}

# answer here
# visualization
dt %>% 
  ggplot(aes(x = Liver)) +
  geom_histogram(breaks = seq(0, 100, 2),
                 fill = "aquamarine4",
                 colour = "white") +
  labs(title = "Distribution of Gene Expression in Liver",
       x = "Expression (TPM)",
       y = "Number of Genes")

# count the number of excluded genes --- 594 genes
dt %>% 
  filter(Liver > 100) %>% 
  nrow()

# Check in lung
dt %>% 
  ggplot(aes(x = Lung)) +
  geom_histogram(breaks = seq(0, 100, 2),
                 fill = "coral3",
                 colour = "white") +
  labs(title = "Distribution of Gene Expression in Lung",
       x = "Expression (TPM)",
       y = "Number of Genes")


### The distribution is strongly right-skewed in liver. Most genes have low or zero expression in the liver, while a small number of genes are highly expressed. This pattern is not specific to the liver because the distribution in the lung is also right-skewed.
```

------------------------------------------------------------------------

### Q2. Density plot

Going back to the liver, now represent the same data with a density plot
rather than a histogram. Do not restrict the X axis. Within the same
code block, create a second plot where you remove the custom bin breaks,
but transform the expression values using `log10()`. First add a
pseudo-count of 1 to the expression values, as the log of 0 is negative
infinity.

What does the log-transformed version show you that the first plot does
not? And where on the log-transformed plot do the genes with 0 TPM end
up?

```{r q2}

# answer here
# density plot
dt %>% 
  ggplot(aes(x = Liver)) +
  geom_density(colour = "aquamarine4") +
  labs(title = "Density of Gene Expression in Liver",
       x = "Expression (TPM)",
       y = "Density")

# density plot
dt %>% 
  ggplot(aes(x = log10(Liver+1))) +
  geom_density(colour = "steelblue") +
  labs(title = "Log-transformed Gene Expression in Liver",
       x = "Expression(log10(TPM + 1))",
       y = "Density")


### The original plot is compressed by a small number of extremely highly expressed genes. The log-transformed plot makes the distribution of low- and moderately-expressed genes easier to see. Genes with 0 TPM appear at log10(0 + 1) = 0.
```

------------------------------------------------------------------------

### Q3. Scatter plot

Make a scatter plot comparing the expression of genes in the `Liver` to
the same genes in the `Heart - Left Ventricle`. Add a pseudocount to
each and log10 transform them. Adjust the point size to 1 and the alpha
(opacity) to 0.2. Add a dashed red line along the diagonal (y = x).

What does a point far above the diagonal mean, biologically?

```{r q3}

# answer here
dt %>% 
  ggplot(aes(x = log10(Liver + 1), 
             y = log10(`Heart - Left Ventricle` + 1))) +
  geom_point(size = 1,
             alpha = 0.2,
             colour = "darkslategrey") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(title = "Compare the Expression of Genes (Liver v.s. Heart-Left Ventricle",
      x = "Liver Expression (log10(TPM + 1))",
      y = "Heart-Left Ventrivle Expression (log10(TPM + 1))")

### A point far above the diagonal represents a gene that is expressed much more highly in the heart than in the liver.
```

------------------------------------------------------------------------

### Q4. Colored points on scatter plot

Color the previous scatter plot points based on whether the genes are
mitochondrial vs. nuclear encoded, using the `is_mitochondrial` column
you made above.

There are only a few dozen mitochondrial genes among tens of thousands
of nuclear ones, so the colored points end up buried. Try plotting them
as a second `geom_point()` layer on top of the others, using the `data`
argument to give that layer just the mitochondrial genes:
`geom_point(data = filter(dt, is_mitochondrial), ...)`.

Where do the mitochondrial genes sit relative to the diagonal, and does
that fit what you found in the last assignment?

```{r q4}

# answer here
# no
dt %>% 
  ggplot(aes(x = log10(Liver + 1), 
             y = log10(`Heart - Left Ventricle` + 1))) +
  geom_point(size = 1,
             alpha = 0.2,
             color = "darkslategrey") +
  geom_point(data = filter(dt, is_mitochondrial),
             size = 1,
             shape = 2,
             alpha = 0.7,
             color = "orange") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  labs(title = "Compare the Expression of Genes (Liver v.s. Heart-Left Ventricle",
       x = "Liver Expression (log10(TPM + 1))",
       y = "Heart-Left Ventrivle Expression (log10(TPM + 1))")

### All mitochondrial genes are above the diagonal, showing that they have higher expression in the heart than in the liver. This agrees with the previous finding because the heart requires large amounts of energy.
```

------------------------------------------------------------------------

### Q5. Facets

Repeat question 1 (histogram), but faceting such that you create
separate panels (arranged horizontally) for several tissues of interest:

```{r focal-tissues}
focal_tissues <- c("Liver", "Testis", "Whole Blood", "Heart - Left Ventricle", "Brain - Cortex")
```

This requires first converting the data into "long" format with
`pivot_longer()`:

```{r pivot-long}
dt_long <- dt %>%
  select("Description", "is_mitochondrial", all_of(focal_tissues)) %>%
  pivot_longer(cols = all_of(focal_tissues), names_to = "Tissue", values_to = "TPM")
```

After creating the 5-panel plot, also split into two rows that separate
genes that are versus are not encoded by the mitochondrial genome (10
panels). Use the argument `scales = "free_y"` to deal with the fact that
there are far fewer mitochondrial-encoded than nuclear-encoded genes.

`scales = "free_y"` means the panels can no longer be compared by eye.
Write the one-sentence note you would put in a figure caption to keep a
reader from being misled.

```{r q5, fig.width=10, fig.height=4}

# your code for 5-panel plot
dt_long %>% 
  ggplot(aes(x = TPM, fill = Tissue)) +
  geom_histogram(breaks = seq(0, 100, 2),
                 colour = "white") +
  facet_wrap(~Tissue, nrow = 1) +
  labs(x = "Expression (TPM)",
       y = "Number of genes") +
  theme(legend.position = "none")
# your code for 10-panel plot
dt_long <- dt_long %>%
  mutate(
    gene_origin = if_else(
      is_mitochondrial,
      "Mitochondrial-encoded",
      "Nuclear-encoded"
    )
  )
dt_long %>% 
  ggplot(aes(x = TPM, fill = Tissue)) +
  geom_histogram(breaks = seq(0, 100, 2),
                 colour = "white") +
  facet_grid(gene_origin ~ Tissue, scales = "free_y") +
  labs(x = "Expression (TPM)",
       y = "Number of genes") +
  theme(legend.position = "none")
# caption note: Histograms are faceted by tissue and gene genomic origin; because y-axis scales are free across panels, bar heights should not be compared directly between mitochondrial-encoded and nuclear-encoded genes.
```

------------------------------------------------------------------------

Optional: Have fun and explore these data! Create your own plots and
interpret them.

```{r optional}

# answer here
```
