# Linear Regression

**Deadline:** Friday, October 9\
**Resubmission deadline:** Friday, October 30

## Assignment Overview

The goal of today's lab is to use linear regression and related statistical methods to investigate the relationship between paternal age, maternal age, and the number of de novo mutations (DNMs) in a proband (offspring). Today's assignment will build familiarity with manipulating tabular datasets containing mixed data types using the **tidyverse** in **R**. Specifically, you will import a table of de novo mutations and manipulate it to calculate the number of maternal and paternal DNMs per individual. You will then fit and interpret simple and multiple linear regression models with **stats::lm**, compare nested models with **stats::anova**, and tidy results with **broom**.

This assignment is an R Markdown notebook. Write your code in the empty code chunks and your answers where you see *Your answer:*.

## Data

Data are taken from [Halldorsson, B. V. et al. (2019). Characterizing mutagenic effects of recombination through a sequence-level genetic map. *Science*, 363(6425)](https://science.sciencemag.org/content/363/6425/eaau1043.abstract).

Read the abstract from the above paper to understand the context of the datasets you will be using. The data you need for this assignment are available from Dropbox at:

1.  Information about the number and parental origin of each de novo mutation detected in a proband (offspring): [aau1043_dnm.csv](https://www.dropbox.com/scl/fi/6e28a3dow872fi02cp537/aau1043_dnm.csv?rlkey=l3gs7fb6igff4el5ov6wg96ai&dl=0)
2.  Ages of the parents of each proband: [aau1043_parental_age.csv](https://www.dropbox.com/scl/fi/sjrq1x1g30h0j10ktxysi/aau1043_parental_age.csv?rlkey=e9g3m9iq4tfsm9ski7w4vb0bf&dl=0)

You may copy these into your submission directory (and add to your `.gitignore`).

Before beginning the assignment, take a quick look at both files (e.g., with `less -S` in Unix) to confirm their structure.

------------------------------------------------------------------------

## Getting started (R packages)

Load the tidyverse and broom packages.

```{r}
library(tidyverse)
library(broom)
```

------------------------------------------------------------------------

## Exercises

### Exercise 1: Wrangle the data

#### **Step 1.1 — Load DNMs**

Load `aau1043_dnm.csv` into a tibble.

```{r}

dnm <- read_csv("aau1043_dnm.csv")
```

#### **Step 1.2 — Count DNMs by parental origin per proband**

Create a **per-proband** summary with counts of maternally and paternally inherited DNMs. Ignore DNMs without a specified parent of origin. Name the count columns `maternal_dnm` and `paternal_dnm`, as later steps use these names.

```{r}
dnm_count <- dnm %>% 
  filter(Phase_combined %in% c("mother", "father")) %>% 
  count(Proband_id, Phase_combined, name = "dmn_count") %>%
  pivot_wider(names_from = Phase_combined, 
              values_from = dmn_count, 
              values_fill = 0) %>% 
  rename(maternal_dnm = mother,
         paternal_dnm = father)
```

#### **Step 1.3 — Load parental ages**

Load `aau1043_parental_age.csv`.

```{r}
parental_age <- read_csv("aau1043_parental_age.csv")
```

#### **Step 1.4 — Merge counts with ages**

Join the two tibbles by proband ID. Name the result `merged`, as later steps use this name.

```{r}
merged <- inner_join(dnm_count, parental_age, by = "Proband_id")
```

------------------------------------------------------------------------

### Exercise 2: Fit and interpret linear regression models with R

Use your merged data frame for the following. All plots should be clearly labeled and easily interpretable.

#### **Step 2.1 — Visualize relationships**

**2.1.1** Create a scatter plot of the count of maternal DNMs vs. maternal age → save as `ex2_a.png`

```{r}
p1 <- merged %>% 
  ggplot(aes(y = maternal_dnm, 
             x = Mother_age)) +
  geom_point(alpha = 0.6,
             colour = "brown2") +
  labs(y = "Count of maternal DNMs",
       x = "Maternal age")
p1
ggsave("ex2_a.png", plot = p1)
```

**2.1.2** Create a scatter plot of the count of paternal DNMs vs. paternal age → save as `ex2_b.png`

```{r}
p2 <- merged %>% 
  ggplot(aes(y = paternal_dnm, 
             x = Father_age)) +
  geom_point(alpha = 0.6,
             colour = "deepskyblue3") +
  labs(y = "Count of paternal DNMs",
       x = "Paternal age")
p2
ggsave("ex2_b.png", plot = p2)
```

**2.1.3** Create a scatter plot of paternal age vs. maternal age → save as `ex2_c.png`

```{r}
p3 <- merged %>% 
  ggplot(aes(x = Father_age,
             y = Mother_age)) +
  geom_point(alpha = 0.6,
             colour = "darkorchid") +
  labs(x = "Paternal age",
       y = "Maternal age")
p3
ggsave("ex2_c.png", plot = p3)
```

#### **Step 2.2 — OLS: maternal age vs. maternal DNMs**

Fit a simple linear regression model relating maternal age to the number of maternal de novo mutations.

```{r}
m1 <- lm(data = merged,
         maternal_dnm ~ Mother_age)
summary(m1)
```

Answer the following questions:

1.  What is the "size" (i.e., slope) of this relationship? Interpret the slope in plain language. Does it match your plot?

    *Your answer: The slope is 0.378. For each additional year of maternal age, the expected number of maternal DNMs increases by 0.378 mutations.*

2.  Is the relationship significant? How do you know? Explain the p-value in plain but precise language.

    \*Your answer: *The relationship is statistically significant because the p-value is less than* $2 \times 10^{-16}$*, which is far below the conventional significance threshold of 0.05. Assuming that maternal age has no true association with the number of maternal DNMs, the probability of observing a relationship at least this strong by random chance is less than* $2 \times 10^{-16}$*.*

#### **Step 2.3 — OLS: paternal age vs. paternal DNMs**

Repeat the step above but for paternal age vs. paternal DNMs.

```{r}
m2 <- lm(data = merged,
         paternal_dnm ~ Father_age)
summary(m2)
```

Answer the following questions:

1.  What is the "size" (i.e., slope) of this relationship? Interpret the slope in plain language. Does it match your plot?

    *Your answer: The slope is 1.354. For each additional year of paternal age, the expected number of paternal DNMs increases by 1.354 mutations.*

2.  Is the relationship significant? How do you know? Explain the p-value in plain but precise language.

    *Your answer: The relationship is statistically significant because the p-value is less than* $2 \times 10^{-16}$*, which is far below the conventional significance threshold of 0.05. Assuming that paternal age has no true association with the number of paternal DNMs, the probability of observing a relationship at least this strong by random chance is less than* $2 \times 10^{-16}$*.*

#### 

#### **Step 2.4 — Predict for a 50.5-year-old father**

Use the paternal regression model to predict the expected number of paternal DNMs for a father of age 50.5. You are welcome to do this manually or using a built-in function, but show your work in the code chunk.

```{r}
new_father <- tibble(Father_age = 50.5)
predict(m2, newdata = new_father)
```

*Your answer: A father of age 50.5 is expected to have 78.7 paternal DNMs*

#### **Step 2.5 — Multiple regression: correlated predictors**

Maternal DNMs arose in the mother's germline, so there is no obvious reason that the father's age should affect them.

**2.5.1** Fit a simple linear regression model relating **paternal** age to the number of **maternal** DNMs.

```{r}
m3 <- lm(data = merged, maternal_dnm ~ Father_age)
summary(m3)
```

1.  Is paternal age a significant predictor of maternal DNMs on its own? Using your plot from 2.1.3, explain why this might happen.

    *Your answer: Paternal age is a significant predictor of maternal DNMs on its own (*$p = 1.37 \times 10^{-13}$*). However, this does not necessarily mean that paternal age directly affects maternal DNMs. The paternal-age versus maternal-age plot shows a positive correlation between the parents' ages, so paternal age may act as a proxy for maternal age, which is the biologically relevant predictor of maternal DNMs.*

**2.5.2** Fit a multiple linear regression model with both maternal age and paternal age as predictors of maternal DNMs, just as we added year as a covariate in the penguin example.

```{r}
m4 <- lm(data = merged, maternal_dnm ~ Mother_age + Father_age)
summary(m4)
```

2.  What happened to the paternal age coefficient and its p-value?

    *Your answer: After including maternal age in the model, the paternal age coefficient became negative (-0.06665) and was no longer statistically significant (p = 0.18773). This suggests that the apparent relationship between paternal age and maternal DNMs was largely due to the correlation between paternal and maternal age.*

3.  Interpret the maternal age coefficient in plain language. How does its meaning differ from the slope you estimated in Step 2.2? (Hint: what is being held constant?)

    *Your answer: Holding paternal age constant, each additional year of maternal age is associated with an expected increase of 0.44498 maternal DNMs. Unlike the slope in Step 2.2, which estimated the unadjusted relationship between maternal age and maternal DNMs, this coefficient estimates the relationship after controlling for paternal age.*

**2.5.3** Use `anova()` to compare your two-predictor model from 2.5.2 to your model from Step 2.2 (maternal age only).

> **Hint:** In the live coding, we used `anova()` to test whether island improved a model that already contained sex and year. The same approach works for any pair of models where the smaller model is the larger model with some terms removed (these are called "nested" models). Here, the smaller model is the one that leaves out paternal age.

```{r}
anova(m4, m1)
```

4.  Does adding paternal age improve the model? Compare the p-value from `anova()` to the p-value for paternal age in the `summary()` of the two-predictor model. Why are they the same here, when for island we needed `anova()` to get a single answer? (Hint: how many coefficients did island add to the model, and how many does paternal age add?)

    *Your answer: Adding paternal age does not significantly improve the model because the ANOVA p-value is 0.1877, which is greater than* $0.05$*. This is the same as the paternal age p-value in the two-predictor model summary (p = 0.18773). The p-values are the same because paternal age adds only one coefficient to the model, so the ANOVA F-test and the t-test for that coefficient test the same hypothesis. In contrast, island had multiple levels and therefore added multiple coefficients. ANOVA was needed to test their joint contribution to the model.*

#### **Step 2.6 — Compare distributions of maternal vs. paternal DNMs**

Plot both distributions on the **same axes** as semi-transparent histograms; save as `ex2_d.png`.

```{r}
dnm_long <- merged %>% 
  select(maternal_dnm, paternal_dnm) %>% 
  pivot_longer(
    cols = c(maternal_dnm, paternal_dnm),
    names_to = "parent",
    values_to = "dnm_count"
  )

p_d <- ggplot(dnm_long, aes(x = dnm_count, fill = parent)) +
  geom_histogram(bins = 30, alpha = 0.5, position = "identity") +
  labs(
    title = "Maternal and paternal DNMs distributions",
    x = "Number of DNMs",
    y = "Number of probands",
    fill = "Parent"
  ) +
  theme_classic()

p_d
ggsave("ex2_d.png", p_d, width = 7, height = 5)
```

#### **Step 2.7 — Statistical test: maternal vs. paternal DNMs per proband**

We have **paired** observations per proband (maternal vs. paternal). The paired t-test assumes that the within-pair differences are approximately normally distributed.

**2.7.1** Apply a paired t-test in R using `t.test(merged$maternal_dnm, merged$paternal_dnm, paired = TRUE)`.

```{r}
t.test(merged$maternal_dnm, merged$paternal_dnm, paired = TRUE)
```

1.  What is the "size" of this relationship (i.e., the average difference in counts of maternal and paternal DNMs)? Interpret the difference in plain language. Does it match your plot?

    *Your answer: The mean difference, calculated as maternal DNMs minus paternal DNMs, is -39.23485. On average, each proband has approximately 39.23 fewer maternal DNMs than paternal DNMs. This matches the plot because the paternal DNM distribution is shifted toward higher values than the maternal DNM distribution.*

2.  Is the relationship significant? How do you know? Explain the p-value in plain but precise language.

    *Your answer: The difference is statistically significant because the p-value is less than* $2.2 \times 10^{-16}$*, which is far below 0.05. Assuming that the true average within-proband difference between maternal and paternal DNM counts is zero, the probability of observing a difference at least or large by random chance is less than* $2.2 \times 10^{-16}$*.*

**2.7.2** The paired t-test is equivalent to using the difference between the maternal and paternal DNM counts per proband as the response variable and fitting a model with only an intercept term (indicated with `1` on the right side of the model formula). Fit this model using `lm()`.

```{r}
merged <- merged %>% 
  mutate(dnm_difference = maternal_dnm - paternal_dnm)

difference_model <- lm(dnm_difference ~ 1, data = merged)

summary(difference_model)
```

3.  How do the results compare to the paired t-test? How would you interpret the coefficient estimate for the intercept term?

    *Your answer: The intercept-only linear model gives the same result as the paired t-test. The estimated mean difference is -39.2348, and the p-value is less than* $2 \times 10^{-16}$*. The intercept represents the average difference of maternal DNMs and paternal DNMs across all probands. Its negative value means that probands have, on average, approximately 39.23 fewer maternal DNMs than paternal DNMs.*

------------------------------------------------------------------------

### Exercise 3: Explore a new dataset

#### **Step 3.1 — Pick a TidyTuesday dataset**

Choose a dataset from the bottom of the [TidyTuesday README](https://github.com/rfordatascience/tidytuesday) and load it.

```{r}
bird <- read_csv("at-sea_seabird_observation_data_bird_data_by_record_id.csv")
ship <- read_csv("at-sea_seabird_observation_data_ship_data_by_record_id_1.csv")

```

1.  Which dataset did you choose?

    *Your answer:* [Bird Sightings at Sea](https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-04-14/readme.md)

#### **Step 3.2 — Explore and visualize**

Generate figures; save them as `ex3_<something>.png`.

```{r}
# total number of birds in the same observation ID
bird_total <- bird %>%
  group_by(`RECORD ID`) %>%
  summarise(total_count = sum(COUNT, na.rm = TRUE),
            .groups = "drop")
# add info: date, latitude, longitude
bird_location <- bird_total %>%
  left_join(ship %>% select(`RECORD ID`, DATE, LAT, LONG),
            by = "RECORD ID") %>%
  mutate(date = dmy(DATE),
         year = year(date)) %>%
  filter(!is.na(LAT), !is.na(LONG), !is.na(total_count))
# plot total seabird count against latitude
# log1p(total_count) is log(total_count + 1), which reduces the influence of very large counts
p_birds <- ggplot(
  bird_location,
  aes(x = LAT, y = log1p(total_count), colour = LONG)
) +
  geom_point(alpha = 0.3,
             size = 0.6) +
  geom_smooth(method = "lm", colour = "black") +
  theme_classic()

p_birds
ggsave("ex3_birds.png", plot = p_birds)
```

1.  What interesting patterns do you notice?

    *Your answer: The plot shows a weak negative association between latitude and log-transformed total seabird count. Observations at lower latitudes (more negative values, farther south) tend to have slightly higher bird counts. However, the relationship is highly variable. Most observations are concentrated between approximately* $-46^\circ$ *and* $-34^\circ$ *latitude, suggesting that sampling locations were not evenly distributed. The largest counts also occur mainly in this latitude range, while many observations across all locations contain few or no birds. Longitude varies across the observations, indicating that geographic location may also help explain seabird abundance.*

#### **Step 3.3 — Pose and test a linear-model hypothesis**

1.  State a hypothesis that you can test with a linear model.

    *Your answer: I hypothesize that the log-transformed total number of seabirds observed is associated with the latitude and longitude of the observation location.*

Fit a linear model with at least two predictors to test your hypothesis, and evaluate its fit.

```{r}
m_birds <- lm(log1p(total_count) ~ LAT + LONG, data = bird_location)

summary(m_birds)
```

2.  Why did you include each predictor?

    *Your answer: I included latitude and longitude because they describe the geographic location of each seabird observation and may be associated with seabird abundance.*

3.  Report and interpret your results. How well does the model fit?

    *Your answer: Both latitude and longitude were statistically significant predictors of log-transformed total seabird count (*$p < 2 \times 10^{-16}$ *for both predictors). Holding longitude constant, a one-degree increase in latitude was associated with a 0.0448 decrease in* $\log(\text{total count} + 1)$*. Holding latitude constant, a one-degree increase in longitude was associated with a 0.0143 increase in* $\log(\text{total count} + 1)$*. However, the model fit was weak: the adjusted* $R^2$ *was 0.0279, meaning that latitude and longitude together explained only about 2.8% of the variation in seabird counts. Therefore, location is statistically associated with seabird abundance, but most variation is explained by other factors not included in this model.*

------------------------------------------------------------------------

## Submission

- Your completed R Markdown notebook (`.Rmd`), with code in each chunk and your answers to all questions.
- The figures you saved (`ex2_a.png` through `ex2_d.png`, and your `ex3_` figures).

## Grading Rubric (Total = 10 points)

### **Exercise 1 — Wrangle the Data (2 points)**

- Load and inspect DNM data (**0.5 pt**)
- Create per-proband maternal and paternal DNM counts (**0.5 pt**)
- Load and inspect parental age data (**0.5 pt**)
- Join counts with ages into a merged table (**0.5 pt**)

### **Exercise 2 — Fit and Interpret Linear Models (6 points)**

- Step 2.1: Scatter plots for DNMs vs. parental age and paternal vs. maternal age (**1 pt**)
- Step 2.2: Fit and interpret maternal OLS model (**1 pt**)
- Step 2.3: Fit and interpret paternal OLS model (**0.5 pt**)
- Step 2.4: Predict paternal DNMs for age 50.5 (**0.5 pt**)
- Step 2.5: Multiple regression with both parental ages, and nested model comparison with `anova()` (**1.5 pts**)
- Step 2.6: Plot distributions of maternal vs. paternal DNMs (**0.5 pt**)
- Step 2.7: Paired t-test (`t.test` and `lm(diff ~ 1)`) and interpret results (**1 pt**)

### **Exercise 3 — Explore a New Dataset (2 points)**

- Step 3.1: Choose and document TidyTuesday dataset (**0.5 pt**)
- Step 3.2: Produce exploratory figure(s) (**0.5 pt**)
- Step 3.3: Pose and test a multiple linear regression hypothesis and interpret results (**1 pt**)

**Total Points: 10**

------------------------------------------------------------------------
